\bf Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-07-07 paper.qmd
acknowledgements

>>>>

## Acknowledgements {-}

The work of first author Calvi was supported by the ResearchFirst undergraduate research program at Monash University. German Valencia advised on the project, in relation to physics applications.

The full paper, code and data to reproduce the work, are publicly available at https://github.com/uschiLaa/anomaly_ppi.

### 2024-07-07 paper.tex
acknowledgements

\caption{\label{fig-liver}Two projections of simulated example data
corresponding to common liver tests: (a) sample of female patients, (b)
longitudinal test results for a single patient. Red cross indicates
observation is outide the 4-D confidence ellipse. The female patients
tend to have lower ALP and ALT than the normal range. As the patient
aged, the level of ALP increases and ALT decreases.}

@@

The new anomaly index is first applied to the full dataset, such that
the final projection will be affected by averaging distance of points in
many directions, see Fig. Figure~\ref{fig-weather} (a). It provides a
global picture showing where the outlying points differ from the normal
distribution. We can see that the contrast between maximum (atmx) and
minimum (atmn) temperature is primarily used in the direction where
points are most outlying. In the orthogonal direction the variables
precipitation (prc) and number of days with maximum temperature above
25°C (nsm) can be used to separate the clusters 2, 3 and 5. With this
solution the two points in cluster 4 cannot be resolved, they fall
inside the cluster. To better understand what is different for those two
years we separately run the anomaly index for this subset, see Fig.
Figure~\ref{fig-weather} (b). From the result we see that the maximum
temperature is not relevant for this cluster, but that they have low
values for the minimum temperature. We also see that there is still some
averaging, and the preferred solution is not placing either point on the
outside of the projected cluster.

@@

>>>>

\caption{\label{fig-liver}Two projections of simulated example data
corresponding to common liver tests: (a) sample of female patients, (b)
longitudinal test results for a single patient. Red cross indicates
observation is outide the 4-D confidence ellipse. The female patients
tend to have lower ALP and ALT than the normal range. As the patient
ages, the level of ALP increases and ALT decreases.}

@@

The new anomaly index is first applied to the full dataset, such that
the final projection will be affected by averaging distance of points in
many directions, see Fig. Figure~\ref{fig-weather} (a). It provides a
global picture showing where the outlying points differ from the normal
distribution. We can see that the contrast between maximum (atmx) and
minimum (atmn) temperature is primarily used in the direction where
points are most outlying. In the orthogonal direction the variables
precipitation (prc) and number of days with maximum temperature above
25°C (nsm) can be used to separate the clusters 2, 3 and 5. With this
solution the two points in cluster 4 cannot be resolved, they fall
inside the ellipse. To better understand what is different for those two
years we separately run the anomaly index for this subset, see Fig.
Figure~\ref{fig-weather} (b). From the result we see that the maximum
temperature is not relevant for this cluster, but that they have low
values for the minimum temperature. We also see that there is still some
averaging, and the preferred solution is not placing either point on the
outside of the projected ellipse.

@@

\section*{Acknowledgements}\label{acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

The work of first author Calvi was supported by the ResearchFirst
undergraduate research program at Monash University. German Valencia
advised on the project, in relation to physics applications.

The full paper, code and data to reproduce the work, are publicly
available at https://github.com/uschiLaa/anomaly\_ppi.
